Recurrent Neural Networks

### 2017-08-24 deeplearn/presentation/paper_2.tex
presentation

Just like any regular neural network layer, recurrent nodes can be composed.  The composition of two recurrent nodes is done by feeding the output of one node into the input of the other, and concatenating their state space.  Formally, if $f:\R^n\times \R^\ell\times\to \R^m\times \R^\ell$ and $g:\R^n\times \R^\ell\times\to \R^m\times \R^\ell$ are two recurrent nodes, then we have the node $(f;g):\R^m\times\R^\ell\times\R^L\to\R^k\times\R^\ell\times\R^L$ defined by $(f;g)(\x, \s_1, \s_2) = (\y, \s'_1, \s'_2)$ where $f(\x, \s_1) = \x', \s'_1$ and $g(\x', \s_2) = \y, \s'_2$. Graphically, we can represent this by stacking the boxes representing $f$ and $g$ on top of one another

@@

\section{Back Propagation}

@@

There are three main types of recurrent cells which are used to build recurrent neural networks.  Let us describe them, and give some of their basic properties, along with a na\"ive implementation in tensorflow.

@@

\begin{minted}{python}
def basic_rnn_cell(input_tensor, state_tensor):
  A_u = tf.Variable(shape=[N, L])
  B_u = tf.Variable(shape=[L, L])
  A_v = tf.Variable(shape=[N, L])
  B_v = tf.Variable(shape=[L, L])
  output_tensor = tf.relu(tf.matmul(input_tensor, A_u) +
                            tf.matmul(state_tensor, B_u))
  new_state_tensor = tf.tanh(tf.matmul(input_tensor, A_v) + \
                                  tf.matmul(state_tensor, B_v))
  return output_tensor, new_state_tensor
\end{minted}

@@

>>>>

Just like any regular neural network layer, recurrent nodes can be composed.  The composition of two recurrent nodes is done by feeding the output of one node into the input of the other, and concatenating their state space.  Formally, if $f:\R^n\times \R^\ell\times\to \R^m\times \R^\ell$ and $g:\R^n\times \R^\ell\times\to \R^m\times \R^\ell$ are two recurrent nodes, then we have the node $(f;g):\R^m\times\R^\ell\times\R^L\to\R^k\times\R^\ell\times\R^L$ defined by $(f;g)(\x, \s_1, \s_2) = (\y, \s'_1, \s'_2)$ where $f(\x, \s_1) = \x', \s'_1$ and $g(\x', \s_2) = \y, \s'_2$. Graphically, we can represent this by stacking the boxes representing $f$ and $g$ on top of one another:

@@

The whole stack can then be seen as one single recurrent node.

\section{Training}

@@

In this section we descripbe three of the main types of recurrent cells currently in use.  They are the basic recurrent cell, which is a straightforward implementation of a function $f:\R^n\times\R^\ell\to\R^\ell$, then long short term memory cell (LSTM), which is more suitable to model long term dependencies in sequences, and the gated recurrent unit cell, which is a simplification of the LSTM cell, but which performs almost equally well at long term dependencies. We give a formal definition of each oc these cells along with a na\"ive implementation in {\tt tensorflow}. The implementation we provide is only expositional though, and would not work out of the box in an actual production model.
%There are three main types of recurrent cells which are used to build recurrent neural networks.  Let us describe them, and give some of their basic properties, along with a na\"ive implementation in tensorflow.

@@

\begin{minted}{python}
def basic_rnn_cell(input_tensor, state_tensor, output_dimension):
  input_dimension = input_tensor.get_shape()[1]
  state_dimension = input_tensor.get_shape()[1]
  A_u = tf.Variable(shape=[input_dimension, output_dimension])
  B_u = tf.Variable(shape=[state_dimension, output_dimension])
  A_v = tf.Variable(shape=[input_dimension, state_dimension])
  B_v = tf.Variable(shape=[state_dimension, state_dimension])
  output_tensor = tf.relu(tf.matmul(input_tensor, A_u) + \
                            tf.matmul(state_tensor, B_u))
  new_state_tensor = tf.tanh(tf.matmul(input_tensor, A_v) + \
                                  tf.matmul(state_tensor, B_v))
  return output_tensor, new_state_tensor
\end{minted}

A variant of this cell uses the output as a state vector, thus computing every step as a function of the cirrent ouput, and the previous output.

@@

As an example of a many-to-many recurrent neural network, we propose an architecture that generates text in the style of a particular author, or body of text.  The architecture is very simple, and consists of three stacked GRU cells.  This model will also expose some of the challenges in training recurrent networks.  As an example, we trained the model on the text of the Harry Potter series, for a total of 500 epochs.  Every once in a while we paused the training to let it generate a sequence of about 2000 characters.  For training we network was unrolled to 30 characters, and the training was done in batches of 32 strings.  The main challenge for training is that the strings should continue from batch to batch.  For example, consider training a similar network unrolled to 2 characters with a batch size of 3.  Consider the sentence {\tt The quick brown fox jumps over the lazy dog}.  The first step is to cut the string into substrings of length 2:
{\tt |Th|e |qu|ic|k |br|ow|n |fo|x |ju|mp|s |ov|er| t|he| l|az|y |do|g.|}. In a normal batching situation, we would then cut this liet of strings into chunks of length 3, like so
{\tt |(Th|e |qu)|(ic|k |br)|(ow|n |fo)|(x |ju|mp)|(s |ov|er)|( t|he| l)|(az|y |do)|} but this causes a problem.  The first sequence of the first batch is {\tt Th}, which will leave the network in a certain state $s$. This For this state to be updated properly, the first element of the second batch should be {\tt |e |}, and {\em not} {\tt |ic|}.  We must therefore use a different batching strategy. The string will give us 7 batches in total, so we number the subsequences with a batch index from 1 to 7 in order, starting over at 1 when we run out of indices. The expected output, in this case, is the same as the input but shifted by one character. In code, the batching looks like:

\begin{minted}{python}
def rnn_minibatch_sequencer(raw_data, batch_size, sequence_size, nb_epochs):
    data = np.array([x for x in raw_data])
    data_len = data.shape[0]
    # using (data_len-1) because we must provide for the sequence shifted by 1 too
    nb_batches = (data_len - 1) // (batch_size * sequence_size)

    total_num_batches = (data_len * nb_epochs) // batch_size
    #batches_per_epoch = len(data_x) // batch_size

    assert nb_batches > 0, "Not enough data, even for a single batch. Try using a smaller batch_size."
    rounded_data_len = nb_batches * batch_size * sequence_size
    xdata = np.reshape(data[0:rounded_data_len], [batch_size, nb_batches * sequence_size])
    ydata = np.reshape(data[1:rounded_data_len + 1], [batch_size, nb_batches * sequence_size])
    I = 0
    for epoch in range(nb_epochs):
        for batch in range(nb_batches):
            x = xdata[:, batch * sequence_size:(batch + 1) * sequence_size]
            y = ydata[:, batch * sequence_size:(batch + 1) * sequence_size]
            x = np.roll(x, -epoch, axis=0)  # to continue the text from epoch to epoch (do not reset rnn state!)
            y = np.roll(y, -epoch, axis=0)
            I += 1
            yield {'train_x':  x,
                   'train_y':  y,
                   'validate': None,
                   'batch_number':  batch,
                   'epoch_number':  epoch+1,
                   'batch_index':   I,
                   'total_batches': total_num_batches,
                   'total_epochs':  nb_epochs}
\end{minted}
